Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2021-04-14 jss/EpiCompare-paper/EpiCompare-paper.tex
Edits to section 3.  I think it now lines up really well with the figure and gives a good picture of what EpiCompare does!

\pkg{EpiCompare} delivers \textit{time-invariant} analysis by (1) taking
a global, not marginal view of how epidemics move through populations
and (2) by treating full epidemics as filamental
trajectories\sout{ and not points produced by functions of time}
\textcolor{orange}{(this concept will be explained in  explained in Section \ref{sec:r0})}.
The following section aims to highlight the strengths of
\textit{time-invariant} \textcolor{orange}{analysis} and define the
mathematical foundations that \pkg{EpiCompare}'s tools stand upon.

Mathematically, epidemics are complex objects. They can be hard to
assess and compare to one another due to the differences in the
diseases, the location where the outbreak occurs, how the affected
population reacts, and the time
\sout{related}\textcolor{orange}{related} features (including start of
the epidemic, speed of infection and more). Time-invariant analysis
makes different epidemics easier to compare by removing many time
dependent aspects of an epidemic. Instead,
\textcolor{orange}{time-invariant analysis} focuses on
\sout{the overall \textcolor{orange}{shape and direction, a filamental trajctory,} \sout{trajectory of }an epidemic.  \textcolor{orange}{These filamental trajectories,  in turn,} }
\textcolor{orange}{filamental trajectories, which emphasize the number of lives affected.}

@@

\(R_0\) is probably the most famous \sout{time-invariant} numerical
summary of an epidemic and is often associated with the
Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2000}.
\(R_0\) is \sout{a one-number summary of a disease and is }defined as
the expected number of infections caused by a single infector who is
added to a completely susceptible population \citep{anderson1992}.
\textcolor{violet}{This \textcolor{orange}{definition has no mention of time and hence} means that $R_0$ is a time-invariant parameter\textcolor{orange}{. Yet $R_0$ is} estimated with time-\sout{based}\textcolor{orange}{dependent} data, which can make it a [difficult quantity to estimate.]}
\textcolor{violet}{\sout{This definition concerns the \textit{generations} of new infections produced by $R_0$ but says nothing about the scale of time in which these new infections can occur.  However, data informing models are often temporally anchored (e.g. weekly case counts).}}\footnote{\textcolor{violet}{This was a bit unclear including defining a new term.   As a result, $R_0$ can be a difficult quantity to estimate and disentangle from time-based data.  For example,}}\textcolor{orange}{For example, \cite{Gallagher2020} demonstrate how $R_0$ can be sensitive to time-\sout{based}\textcolor{orange}{dependent} parameters such as the beginning and end of an epidemic, two quantities that generally}\textcolor{violet}{hard to define precisely.\sout{do not have precise definitions}}.
To demonstrate the difficulty of discerning \(R_0\) in
\textcolor{violet}{\sout{a}other}\footnote{\textcolor{violet}{I change this so we don't confused readers that we're going show the impact in tools beyond just the estimation itself.}}
time-dependent analysis, we first introduce \citet{Kermack1927}'s SIR
model. This model captures the transitions from one state to the next as
a system of ordinary differential equations, where \(N\) is the total
number of individuals, \(\beta\) is the rate of infection, and
\(\gamma\) is the rate of recovery,

@@

In this section, we present the tools implemented in \pkg{EpiCompare}
and explain how they aid in the data analysis pipeline. In Fig.
\ref{fig:pipeline2}, we illustrate how our package's functions fit into
the data analysis pipeline introduced in Fig. \ref{fig:pipeline}. All
front-facing functions are aimed to be as user-friendly as possible. We
also focus on providing the user ``tidyverse'' style functions, that
encourage piping and also follow clear verb naming schemes
\citep{Wickham2019}. Although users can typically incorporate
\pkg{EpiCompare} into any step in the data analysis process, there are
two primary points of entry. The first point of entry is the very
beginning with pre-processing and visualizing raw data, and the second
point of entry is after modeling and simulation. Figure
\ref{fig:pipeline2} captures these different paths, and we will
highlight\footnote{\textcolor{violet}{[Ben says: we need to make sure we actually do highlight ]}}
both approaches and how to leverage \pkg{EpiCompare} in the subsections
below.

@@

The first step of most data analysis is cleaning the data so it can be
explored. There are multiple ways to collect epidemiological data.
Sometimes individual records are collected, with times of different
states of the epidemic (infection, recovery, etc.) as well as individual
information like network structure, location, and sub-population
information. Other data collections focus on aggregate counts of
individuals in each epidemic state. In fact, usually only the number of
new infections at each time step (e.g.~weekly case counts) is observed.
Compartment totals (amounts of individuals in each state) are then
imputed from those case counts along with other information about the
disease and the population of interest. In \pkg{EpiCompare}, we focus on
understanding the overall impact of an outbreak at the
aggregate/population level, which allows for examination of overall
trends of an epidemic more easily.

In order to help the practitioner examine epidemics from an aggregate/
population lens, we provide a function called
\texttt{agents\_to\_aggregate}. This function transforms data about
individual/agents' initial entry into each state (e.g.~start of
infection, start of recovery, etc.) to an aggregate view of how many
individuals were in a state at a given time. There are often situations
where grouping agents into subpopulations (e.g.~subpopulations defined
by age or sex) can highlight different aggregate level trends. For
example, research by \citet{rvachev1985,anderson1992,worby2015} develop
state-based models that account for subpopulations. In \pkg{EpiCompare},
we facilitate subpopulation analysis by combining
\pkg{dplyr}::\texttt{group\_by} and \texttt{agent\_to\_aggregate} to
provide aggregation at a group level.

The \texttt{agents\_to\_aggregate} function is flexible and can deal
with a wide range of information about each individual. It can,
theoretically, account for infinitely many states. This functionality
allows the practitioner to aggregate information relative to the
standard states (e.g.~``Susceptible'', ``Infectious'', and
``Recovered'') as well as add states (e.g.~``Exposed'', ``iMmune'',
``Hospitalized'')
(CITE\footnote{\textcolor{violet}{[Ben says: this was originally to back up the claim that SIR is the "standard" states - now we might need to have a paper that says that SIR is the standard states and also suggests other states...]}}).
Additionally, \texttt{agents\_to\_aggregate} also permits indicators for
death/exit and birth/entry dates. Overall, the function
\texttt{agents\_to\_aggregate()} is a powerful tool for pre-processing
data.

@@

With raw data, ``getting to know'' our data currently means figuring out
good combinations of visualizations, numerical summaries and subsets. An
expert coder has many ways to successfully explore the data in an
aggregate lens using \texttt{agents\_to\_aggregate}. \pkg{EpiCompare}
also includes tools to rapidly explore data that has three epidemic
states. Building on the tools in \pkg{ggplot2} and \pkg{ggtern}, our
\texttt{geom\_aggregate} provides a rapid way to explore different
subpopulations' experience of an epidemic by combining the ideas behind
\texttt{agents\_to\_aggregate} for three-state models to examine
subpopulation trajectories in a 3d simplex space
\citep{Wickham2016, Hamilton2018}.
\textcolor{orange}{[Shannon says: come back - think about spaces and trajectories?]}\footnote{\textcolor{violet}{[Ben says: what does this note mean?]}}
Visualization tools for three-state models were developed because (1)
SIR models are some of the most common and basic epidemic state-based
models and (2) our simplex representation of these epidemics emphasizes
a ``time-invarance'' representation of the data (for a refresher see
Section \ref{sec:time-invariant}).
\textcolor{orange}{[Shannon says: make sure SIR is defined before.]}

@@

\textcolor{violet}{After getting a good sense of what a past or current epidemice looks like through EDA, the next step is often model fitting and/or simulations. In this step, and the next step (post-processing), we discuss how to include practioners' use of tools to create models and simulations that exist outside of \pkg{EpiCompare}.}
This package does not focus on fitting a model to data, we do provide
some flexible functions for simulation of basic discrete-time
epidemic-state models.
\textcolor{violet}{After estimating or predicting transition rates between states,}
these functions produce individual level
\textcolor{violet}{simulations\sout{information}} and can be
\textcolor{violet}{\sout{naturally}} combined with
\texttt{agents\_to\_aggregate()} to view these simulations through an
aggregate lens. The function \texttt{simulate\_SIR\_agents()} simulates
an SIR epidemic with user inputs for the number of simulations, the
initial number in each state, the infection and recovery parameters
\((\beta, \gamma\)), and the total number of discrete time steps. This
function allows for easy access to SIR model analysis and comparison.
Beyond SIR models, the function \texttt{simulate\_agents()} takes as
input a user-specified transition matrix and other epidemic parameters
to allow the user to create simulations of an outbreak for \textit{any}
number of states and any number of transitions among them. This
flexibility in states can be used to also reflect group-based dynamics.
This allows for users to explore the space of models in an intuitive way
without getting bogged down by too much mathematical detail. For
consistency, we have made output from \texttt{simulate\_agents()} and
\texttt{simulate\_SIR\_agents()} compatible with
\texttt{agents\_to\_aggregate()} so aggregate information may easily be
accessed.

\textcolor{orange}{READ}

\textbf{Post-processing}

\textcolor{violet}{[Ben says: I think we should remind the reader that we care more about simulations, in order to compare fitted models between themselves and the true epidemics.]}

\textcolor{violet}{[Ben says: this replaces the first paragraph below]}

\textcolor{violet}{If the practitioner wishes to compare between models and with their true epidemics, one needs to process their models and simulations.}
\textcolor{violet}{In general, [p}ost-processing of modeling and
simulation consists of making summary statistics, plots, tables, and
other ways to disseminate information to the public.{]}
\textcolor{violet}{The summaries can be very complex, and a}{[}s a
result, a number of epidemic modeling \textcolor{violet}{\proglang{R}}
packages return a special class, specific to their modeling. The special
classes often contain a plethora of information from residuals, model
diagnostics, input parameters, and more. While incredibly useful, these
special classes can be difficult for novice coders to navigate.{]}

\textcolor{violet}{[Ben says: replaced with above paragraph]}
Post-processing of modeling and simulation consists of making summary
statistics, plots, tables, and other ways to disseminate information to
the public. For example, comma separated value files (\texttt{.csv}) are
a standard way to share information within tables. However, model output
is often far more complicated than what a traditional \texttt{.csv}
would allow. \textcolor{orange}{Do we need three sentences about csvs?}
As a result, a number of epidemic modeling packages return a special
class, specific to their modeling. The special classes often contain a
plethora of information from residuals, model diagnostics, input
parameters, and more. While incredibly useful, these special classes can
be difficult for novice coders to navigate.

To this end, we have \textcolor{violet}{provide\sout{adapted}} a series
of \texttt{fortify}-style
\textcolor{violet}{methods\sout{functions}}\footnote{\textcolor{violet}{[Ben says: Not tied to this change.]}},
called \texttt{fortify\_aggregate()} which transform output from
packages like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data frames
which contain information about the total number of individuals in each
state at a given time, for a given simulation. These fortify functions
have output that is consistent with that of
\texttt{agents\_to\_aggregate()}.

\textcolor{violet}{[Ben says: new paragraph]} To utilize simulations
epidemics in later time-invariant analysis we also provide a function to
convert temporally defined epidemics to filamental representations.
Specifically, we provide the function \texttt{filament\_compression} to
convert simulation(s) to filaments as expressed by presenting the
epidemic as a ordering of some common fixed number of points so that
they are equally spaced along the original path in the proportional
state
space\footnote{\textcolor{violet}{[Ben: this should be cleaned up.]}}.

@@

As introduced in Section \ref{sec:beyond-r0-sir} there's a lot of
potential for time-invariant tools to help compare and assess epidemics
and models/simulations. In \pkg{EpiCompare} we provide a set of
comparison and assessment tools for models that extend beyond the
standard performance metrics (e.g.~means squared error, AIC) and focus
on assessing the structural information the models capture. This
approaches work well on models where online one ``cycle'' of the
epidemic has occurred (no recovered individuals have been susceptible
again)\footnote{\textcolor{violet}{[Ben says: Shannon - do you think this is clear / a desirable way to define this - we define it slightly differently in section 2.2.]}}.
Epidemics are complex objects, and we provide tools to create prediction
regions out of simulated epidemics. For three-state epidemic models, we
provide the \texttt{ggplot}/\texttt{ggtern} extension
\texttt{geom\_prediction\_band()} which creates a prediction region
around the top \(1-\alpha\) proportion of the simulations (where the
simulations treated as filaments). In \pkg{EpiCompare} we also provide
these prediction regions for epidemic models with with more than three
states. The functions \texttt{create\_convex\_hull\_structure} and
\texttt{create\_delta\_ball\_structure} create different geometric
representations of prediction regions for any state-based model. For
both of these geometric objects, we provide functions to check if a path
is contained (\texttt{contained}) and the ability to assess the Hassdorf
distance between prediction regions based on simulations from different
model (\texttt{hassdorf\_dist}).

We also provide functions to calculate the extremeness of a true
epidemic compared to simulated epidemics through the equi-distance
filamental trajectory representation as mentioned in Section
\ref{sec:beyond-r0-sir}. Specifically, functions like
\texttt{distance\_psuedo\_density\_function} can calculate a
psuedo-density estimate of the true epidemic relative to simulated ones.
Functions \texttt{distance\_depth\_function} and
\texttt{local\_distance\_depth\_function} provide depth scores that
suggest how geometrically central an epidemic is to simulations.
\textcolor{violet}{[Ben says: should we define these more in this paragraph? Introducted in 2.2. as well and not defined.]}

\textcolor{violet}{[Ben says: Replaced with paragraphs above.]}
Comparison and assessment of model fit or comparisons of one model to
another model can be performed in a variety of ways including mean
square error, AIC, plots, and more. Perhaps the most useful tool
\pkg{EpiCompare} has to offer to the expert, for comparison and
assessment of models, is in its post-processing tools which create a
standard output. It is then a matter of writing a script or function
made for that standard output to assess the results from multiple models
in the way the user
desires\footnote{\textcolor{violet}{[Ben says: Shannon, doesn't this below in the previous section?, I might drop this paragraph except for the introduction sentence - is it really true that the standardization is most valuable?]}}.

@@

With \pkg{EpiCompare}, we can easily obtain the empirical cumulative
incidence function with respect to the measles rash appearance (variable
\code{ERU}) with the following tidy-style function,
\code{agents_to_aggregate}. The function \code{agents_to_aggregate} is a
key component of \pkg{EpiCompare}, allowing the user to easily switch
from an individual-level (i.e.~an agent) view of a disease to an
aggregate level. For example, the below code shows how we can convert
the agent data to a cumulative incidence of the measles rash, in order
to see how the disease spread through the population over time. We can
then compare the cumulative incidence of the rash to the cumulative
incidence of the prodromes, i.e.~the initial symptoms. We do this with
the below code, and a part of the cumulative incidence data output are
shown in Table \ref{tab:cif-rash}. The argument
\code{integer_time_expansion} indicates whether we should include all
time points in the recorded range of the data or only when there is a
change in the incidence.

@@

Moreover, we can look at the outbreaks of the disease by group within
\code{agent_to_aggregate()} or \code{geom_aggregate()}. This allows us
to examine differences among the different groups of individuals. For
example, we show the time invariant outbreak by class level in Figure
\ref{fig:tern-class-data}. Immediately, we see that time invariant
infection curve is different for the pre-school class compared to the
1st class. In the 1st class, we see about 95\% of the class become
infected and less than 10\% of them having recovered, which is
indicative of a super-spreading event. This suspicion is further
confirmed in that 26 of the 30 1st class students have been reportedly
infected by the same individual.

@@

However, both models are not a good fit to the filamental path as
opposed to the individual points in \((S, I, R)\)-space. This can be
captures with the set of simulations both models predict, which all
generally have a single defined peak of infection whereas the data
certainly looks like it has two distinct peaks, likely caused by our
assumed super-spreader event. This observation is backed up by the below
analysis that demonstrates that the estimated psuedo-density of the
observed epidemic (relative to the simulations from either model) is
much less likely then \textbf{any} of the simulations (reported in Table
\ref{tab:hags-extreme}. In conclusion, \pkg{EpiCompare} makes it clear
that, at a glance, 1) the EpiModel network model is a better fit than
the Simple SIR model, and 2) the fit is only good at the individual
point level as opposed to the epidemic path level.

>>>>

\pkg{EpiCompare} delivers \textit{time-invariant} analysis by (1) taking
a global, not marginal view of how epidemics move through populations
and (2) by treating full epidemics as filamental trajectories. The
following section aims to highlight the strengths of
\textit{time-invariant} \textcolor{orange}{analysis} and define the
mathematical foundations that \pkg{EpiCompare}'s tools stand upon.

Mathematically, epidemics are complex objects. They can be hard to
assess and compare to one another due to the differences in the
diseases, the location where the outbreak occurs, how the affected
population reacts, and the time
\sout{related}\textcolor{orange}{related} features (including start of
the epidemic, speed of infection and more). Time-invariant analysis
makes different epidemics easier to compare by removing many time
dependent aspects of an epidemic. Instead,
\textcolor{orange}{time-invariant analysis} focuses
\textcolor{orange}{on the global pattern of an epidemic, via filamental trajectories,  and emphasizes the number of lives affected.}
\textcolor{violet}{[Ben wants to try this sentence again.]}

@@

\(R_0\) is probably the most famous \sout{time-invariant} numerical
summary of an epidemic and is often associated with the
Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2000}.
\(R_0\) is \sout{a one-number summary of a disease and is }defined as
the expected number of infections caused by a single infector who is
added to a completely susceptible population \citep{anderson1992}.
\textcolor{violet}{This \textcolor{orange}{definition has no mention of time and hence} means that $R_0$ is a time-invariant parameter\textcolor{orange}{. Yet $R_0$ is} estimated with time-\sout{based}\textcolor{orange}{dependent} data, which can make it a difficult quantity to estimate.}
\textcolor{orange}{For example, \cite{Gallagher2020} demonstrate how $R_0$ can be sensitive to time-\sout{based}\textcolor{orange}{dependent} parameters such as the beginning and end of an epidemic, two quantities that generally are}\textcolor{violet}{hard to define precisely.\sout{do not have precise definitions}}.
To demonstrate the difficulty of discerning \(R_0\) in
\textcolor{violet}{\sout{a}other}\footnote{\textcolor{violet}{I change this so we don't confused readers that we're going show the impact in tools beyond just the estimation itself.}}
time-dependent analysis, we first introduce \citet{Kermack1927}'s SIR
model. This model captures the transitions from one state to the next as
a system of ordinary differential equations, where \(N\) is the total
number of individuals, \(\beta\) is the rate of infection, and
\(\gamma\) is the rate of recovery,

@@

In this section, we present the tools implemented in \pkg{EpiCompare}
and explain how they aid in the data analysis pipeline. In Fig.
\ref{fig:pipeline2}, we illustrate how our package's functions fit into
the data analysis pipeline introduced in Fig. \ref{fig:pipeline}. All
front-facing functions are aimed to be as user-friendly as possible. We
also focus on providing the user ``tidyverse'' style functions, that
encourage piping objects from one function to the next and follow clear
``verb'' naming schemes \citep{Wickham2019}. Although users can
\sout{typically} incorporate \pkg{EpiCompare} into any step in the data
analysis \sout{process}\textcolor{orange}{pipeline}, there are two
primary points of entry. The first point of entry is the very beginning
with pre-processing and visualizing raw data, and the second point of
entry is after modeling and simulation. Figure \ref{fig:pipeline2}
captures these different paths, and we
highlight\footnote{\textcolor{violet}{[Ben says: we need to make sure we actually do highlight ]}}
both approaches and how to leverage \pkg{EpiCompare} in the subsections
below.

@@

The first step of most data analysis is ``cleaning'' the data
\textcolor{orange}{to a format that is friendly for both computers and programmers}
so it can be explored.
\sout{There are multiple ways to collect epidemiological data.}
\textcolor{orange}{Epidemiological data are collected in many different formats.}
Sometimes individual records are collected, with times of different
states of the epidemic (infection, recovery, etc.) as well as individual
information like network structure, location, and sub-population
information. Other data collections focus on aggregate counts of
individuals in each epidemic state. In fact,
\sout{usually}\textcolor{orange}{many times} only the number of new
infections at each time step (e.g.~weekly case counts) is observed.
Compartment totals (amounts of individuals in each state) are then
imputed from those case counts\sout{ along with}\textcolor{orange}{and}
other information about the disease and the population of interest. In
\pkg{EpiCompare}, we focus on understanding the overall impact of an
outbreak at the aggregate/population level, which allows for
\textcolor{orange}{streamlined} examination of overall trends of an
epidemic.\sout{ more easily.}

\sout{In order t},To help the practitioner examine epidemics from an
aggregate/ population lens, we provide a function called
\texttt{agents\_to\_aggregate()}. This function transforms data about
individual/agents' initial entry into each state (e.g.~start of
infection, start of recovery, etc.) to an aggregate view of how many
individuals were in a state at a given time. There are often situations
where grouping agents into subpopulations
(e.g.~sub\sout{populations}\textcolor{orange}{ groups} defined by age or
sex) can highlight different aggregate level trends. For example,
research by \citet{rvachev1985,anderson1992,worby2015} develop
state-based models that account for
\textcolor{orange}{differing disease dynamics in different}
subpopulations. In \pkg{EpiCompare}, we facilitate subpopulation
analysis by combining \textcolor{orange}{the function}
\pkg{dplyr}::\texttt{group\_by()} and \texttt{agent\_to\_aggregate()} to
provide aggregation \sout{at a}\textcolor{orange}{by} group level.

The \texttt{agents\_to\_aggregate()} function is flexible and can deal
with a wide range of information about each individual. \sout{It can,}
Theoretically, account for infinitely many states. This functionality
allows the practitioner to aggregate information relative to
\sout{the standard}\textcolor{orange}{common} states
(e.g.~``Susceptible'', ``Infectious'', and ``Recovered'') as well as
\sout{add}\textcolor{orange}{more complex} states (e.g.~``Exposed'',
``iMmune'', ``Hospitalized'')
(CITE\footnote{\textcolor{violet}{[Ben says: this was originally to back up the claim that SIR is the "standard" states - now we might need to have a paper that says that SIR is the standard states and also suggests other states...]}\textcolor{orange}{Shannon says: I changed 'standard' to 'common' so maybe that's a less loaded word.  Otherwise I think I'd just cite anderson and may 1992}}).
Additionally, \texttt{agents\_to\_aggregate()} \sout{also} permits
indicators for death/exit and birth/entry dates. Overall, this function
is a powerful tool for pre-processing data,
\textcolor{orange}{and it lowers the barrier for entry into data analysis for less experienced practicioners.}\footnote{\textcolor{orange}{Shannon says: I added this in theme with 'highlighting a point of entry'}}

@@

\sout{With raw data, }``Getting to know'' \sout{our} data currently
means figuring out \sout{good}\textcolor{orange}{useful} combinations of
visualizations, numerical summaries and
\sout{subsets}\textcolor{orange}{groupings}. An expert coder has many
ways to successfully explore the data in an aggregate lens using
\texttt{agents\_to\_aggregate()}.
\textcolor{orange}{For less experienced coders,} \pkg{EpiCompare} also
includes tools to rapidly explore data that has three epidemic states.
Building on the tools in \pkg{ggplot2} and \pkg{ggtern},
\sout{our}\pkg{EpiCompare}`s \texttt{geom\_aggregate()} provides a rapid
way to explore different subpopulations' experiences of an epidemic
\citep{Wickham2016, Hamilton2018}. \textbackslash sout\{By combining the
ideas behind agents\_to\_aggregate for three-state models to examine
subpopulation trajectories in a 3d simplex space\}
\citep{Wickham2016, Hamilton2018}\footnote{\textcolor{orange}{I got rid of the end of this sentence because I think it was putting multiple ideas in one sentence.}}.
Visualization tools for three-state models were developed because (1)
SIR models are some of the most common and basic epidemic state-based
models and (2) our \textcolor{orange}{3d} simplex representation of
these epidemics emphasizes a ``time-invarance'' representation of the
data (for a refresher see Section \ref{sec:time-invariant}).

@@

. The special classes often contain a plethora of information from
residuals, model diagnostics, input parameters, and more. While
incredibly useful, these special classes can be difficult for novice
coders to navigate.{]}

{[}\textcolor{violet}{[Ben says: replaced with above paragraph]}
Post-processing of modeling and simulation consists of making summary
statistics, plots, tables, and other ways to disseminate information to
the public. For example, comma separated value files (\texttt{.csv}) are
a standard way to share information within tables. However, model output
is often far more complicated than what a traditional \texttt{.csv}
would allow. \textcolor{orange}{Do we need three sentences about csvs?}
As a result, a number of epidemic modeling packages return a special
class, specific to their modeling. The special classes often contain a
plethora of information from residuals, model diagnostics, input
parameters, and more. While incredibly useful, these special classes can
be difficult for novice coders to navigate. {]}

To this end, \pkg{EpiCompare} \textcolor{violet}{provides\sout{adapted}}
a series of \texttt{fortify}-style
\textcolor{violet}{methods\sout{functions}}\footnote{\textcolor{violet}{[Ben says: Not tied to this change.]}},
called \texttt{fortify\_aggregate()} which transform output from
\textcolor{orange}{infectious disease modeling and simulation} packages
like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data frames which
contain information about the total number of individuals in each state
at a given time, for a given simulation. These fortify functions have
output that is consistent with that of \texttt{agents\_to\_aggregate()}.

\textcolor{violet}{[Ben says: new paragraph]} To utilize simulations of
epidemics in later time-invariant analysis we also provide a function to
convert temporally defined epidemics to filamental representations.
Specifically, we provide the function \texttt{filament\_compression()}
to convert simulation(s) to filaments as expressed by presenting the
epidemic as a ordering of some common fixed number of points so that
they are equally spaced along the original path in the proportional
state
space\footnote{\textcolor{violet}{[Ben: this should be cleaned up.  I also 'highlighted' why this makes epicompare a good entry point]}}

\textcolor{orange}{Shannon's attempt at above paragraph is below..  I tried to more smoothly connect to previous paragraph and then kinda copped out at trying to reword the filament description by just referring back to the previous section.}

{[}\pkg{EpiCompare} also provides a tool to convert time-dependent
epidemic simulations into their time-invariant filamental
representations. Specifically, \texttt{filament\_compression()} converts
simulations to filaments (see Section \ref{sec:time-invariant}){]} so
practitioners can view model simulations and results through a
time-invariant lens. These tools were developed to provide another
natural entry point into the \pkg{EpiCompare} data analysis pipeline for
situations where modeling and analysis is already completed and
practitioners are looking for simple and transparent tools to help
understand and disseminate results.{]}

@@

\textcolor{orange}{Finally, \pkg{EpiCompare} can be used the last step of the data analysis pipeline with its comparison and assessment utility functions.}
As introduced in Section \ref{sec:beyond-r0-sir} there's a
\sout{lot of}\textcolor{orange}{much} potential for time-invariant tools
to help compare and assess epidemics and models/simulations. In
\pkg{EpiCompare} we provide a set of comparison and assessment tools for
models that extend beyond the standard performance metrics (e.g.~mean
squared error or AIC) and focus on assessing the structural information
the models capture. This approaches work well on models where
\sout{online} one ``cycle'' of the epidemic has occurred (no recovered
individuals have been susceptible
again)\footnote{\textcolor{violet}{[Ben says: Shannon - do you think this is clear / a desirable way to define this - we define it slightly differently in section 2.2.]}\textcolor{orange}{Shannon says I don't think I understand what you are trying to say there.  We should chat about it.}}.
Epidemics are complex objects, and we provide tools to create prediction
regions
\textcolor{orange}{with differing desired characteristics}\footnote{\textcolor{orange}{Shannon says: I want to connect epidemics being complex to the fact that prediction is hard and not all prediction regions tell you the same thing}}
from simulated epidemics. For three-state epidemic models, we provide
the \texttt{ggplot}/\texttt{ggtern} extension
\texttt{geom\_prediction\_band()} which creates a prediction region
around the top \(1-\alpha\) proportion
{[}\textcolor{orange}{good place for Ben paper cite?}{]} of the
simulations (where the simulations treated as filaments). In
\pkg{EpiCompare} we also provide these prediction regions for epidemic
models with with more than three states. The functions
\texttt{create\_convex\_hull\_structure()} and
\texttt{create\_delta\_ball\_structure()} create different geometric
representations of prediction regions for any state-based model. For
both of these geometric objects, we provide functions to check if a path
is contained (\texttt{contained()}) and the ability to assess the
Hausdorff distance between prediction regions based on simulations from
different model (\texttt{hassdorf\_dist()}).
\textcolor{orange}{I don't think so, best left to the vignette and data example}

We also provide functions to calculate the ``extremeness'' of a true
epidemic {[}\textcolor{orange}{trajectory?}{]} compared to simulated
epidemics via the equi-distance filamental trajectory representation as
mentioned in Section \ref{sec:beyond-r0-sir}. Specifically, functions
like \texttt{distance\_pseudo\_density\_function()} can calculate a
pseudo-density estimate of the true epidemic relative to simulated ones.
Functions \texttt{distance\_depth\_function()} and
\texttt{local\_distance\_depth\_function()} provide depth scores that
suggest how geometrically central an epidemic is to simulations.
\textcolor{violet}{[Ben says: should we define these more in this paragraph? Introducted in 2.2. as well and not defined.]}\textcolor{orange}{[I'd throw a citation or two but again limit the detail.]}

\textcolor{violet}{[Ben says: Replaced with paragraphs above.]}
Comparison and assessment of model fit or comparisons of one model to
another model can be performed in a variety of ways including mean
square error, AIC, plots, and more. Perhaps the most useful tool
\pkg{EpiCompare} has to offer to the expert, for comparison and
assessment of models, is in its post-processing tools which create a
standard output. It is then a matter of writing a script or function
made for that standard output to assess the results from multiple models
in the way the user
desires\footnote{\textcolor{violet}{[Ben says: Shannon, doesn't this below in the previous section?, I might drop this paragraph except for the introduction sentence - is it really true that the standardization is most valuable?]}\textcolor{orange}{I agree with you now that this section is nice and filled in.  Let's drop this paragraph.}}.

@@

With \pkg{EpiCompare}, we can easily obtain the empirical cumulative
incidence function with respect to the measles rash appearance (variable
\code{ERU}) with the following tidy-style function,
\code{agents_to_aggregate()}. The function \code{agents_to_aggregate()}
is a key component of \pkg{EpiCompare}, allowing the user to easily
switch from an individual-level (i.e.~an agent) view of a disease to an
aggregate level. For example, the below code shows how we can convert
the agent data to a cumulative incidence of the measles rash, in order
to see how the disease spread through the population over time. We can
then compare the cumulative incidence of the rash to the cumulative
incidence of the prodromes, i.e.~the initial symptoms. We do this with
the below code, and a part of the cumulative incidence data output is
shown in Table \ref{tab:cif-rash}. The argument
\code{integer_time_expansion} indicates whether we should include all
time points in the recorded range of the data or only when there is a
change in the incidence.

@@

Moreover, we can look at the outbreaks of the disease by group within
\code{agent_to_aggregate()} or \code{geom_aggregate()}. This allows us
to examine differences among the different groups of individuals. For
example, we show the time invariant outbreak by class level in Figure
\ref{fig:tern-class-data}. Immediately, we see that time invariant
infection curve is different for the pre-school class compared to the
1st class. In the 1st class, we see about 95\% of the class become
infected and less than 10\% of them having recovered, which may be
indicative of a super-spreading event. This suspicion is further
confirmed in that 26 of the 30 1st class students have been reportedly
infected by the same individual.

@@

However, both models are not a good fit to the filamental path as
opposed to the individual points in \((S, I, R)\)-space. This can be
captured with the set of simulations both models predict (gray lines),
which all generally have a single defined peak of infection whereas the
data certainly looks like it has two distinct peaks, likely caused by
our assumed super-spreader event. This observation is backed up by the
below analysis that demonstrates that the estimated pseudo-density of
the observed epidemic (relative to the simulations from either model) is
much less likely then \textbf{any} of the simulations (reported in Table
\ref{tab:hags-extreme}). In conclusion, \pkg{EpiCompare} makes it clear
that, at a glance, 1) the \pkg{EpiModel} network model is a better fit
than the Simple SIR model, and 2) the fit is only good at the geometric
filamental level as opposed to the epidemic trajectory filamental level.
